---
title: "Control Structures"
author: "Travis Brady"
date: "1/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Control Structures

## General

- **if, else** tests a condition
- **for** execute a loop a fixed number of times
- **while** execute a loop while a condition is true
-**repeat** execute infinite loop
-**break** break loop execution
-**next** skip iteration of a loop
-**return** to exit a function

## If/If-else

```{r, echo = F}
## if(<condition>) {do something}
## else if(<condition>) {do something different}
## else {do something else}
```

```{r}
x <- 12
if(x>0) {
  y <- 10
} else {
  y <-2}
# else-if(x>3) {y <- 12}
```

## For Loop

**seq_along** can be used to create an integer sequence equal to the length of the vector input. Below are equivalent processes for printing individual values in a for loop.

```{r}
x <- c("a","g", "b", "c", "z", "d", "e", "f")
for(i in 1:8) {
  print(x[i])}
for(i in seq_along(x)) {print(x[i])}
for(element in x) print(element)
for(i in 1:8) print(x[i])
```


For loops can be nested (matrix style):

```{r}
m1 <- matrix(1:6, 2, 3)
for(i in seq_len(nrow(m1))) {## index along rows (cannot use seq_along since this has 2 dim)
  for(j in seq_len(ncol(m1))) {## index along number of columns
    print(m1[i,j])}} ## print values
```

While Loops:


```{r}
count <- 0
while(count<=10) {
  print(count)
  count <- count +1} ## make sure condition that stops the loop occurs!
```

```{r}
z <- 5 ## start with 5
while(z >= 3 & z <= 10) { ## set condition of loop, z is between 3 and 10 inclusive
  print(z)
  coin <- rbinom(1, 1, 0.5) ## "one flip of one coin with a 50% probability of being heads or tails"
  if(coin == 1) { 
    z <- z + 1 ## "add one for heads"
  } else { z <- z -1}} ## subtract one for tails
```









