---
title: "R_Programming_Markdown_Notes"
author: "Travis Brady"
date: "12/30/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Changing Directories

**getwd()** will provide you with working directory. You can only read data that is currently in your working directory
dir() lists all files in the current directory. read.csv("") can be used to import and read CSV files, given they are placed in the correct directory.

# Console Input and Evaluation

**<-** is the assignment operator, it can be used anywhere. You can autoprint objects or explicitly print them with the print function. The brackets before the line of code in the result tells you which elements of the vector you are looking at. 

```{r}
x <- 1 ## assign x the value of 1
y <- 1:20 ## assign y the vector 1:20
print(x) ## print x
msg <- "hello" ##assign msg the string "hello"
```

# Objects and Attributes

R has 5 basic (atomic) classes of objects:

- character
- numeric
- integer
- complex
- logical/Boullion (True/False)

Vectors are the most basic

- all objects in a vector are of the same class
- **lists are a key exception**; they contain different classes of objects
- empty vectors can be created with the **vector()** function - first argument is the class, second argument is the length.

## Numbers

- numbers can be intergers or numeric objects: 1 is a numeric object, 1L gives the integer
- **inf** is infinity
- NaN (not a number) is for undefined values (0/0 for example)


## Attributes

- names and dimnames
- dimensions
- class
- length
- other user-defined attributes
- attributes can be provided using **attributes()** function

# Vectors, Lists and Matrices

## Vectors

The **c()** function creates vectors that are numeric, logical, characters, or integers. Combining different classes will not return an error, but R will "coerce" elements of the vector to have uniforminty. Preferentially made into characters or numerics.

```{r}
a <- c(0, 0.3, 1, 0.5)
b <- c(TRUE, FALSE)
c <- c(T,F)
d <- c("a", "b", "c")
e <- c(10:32)
f <- c(1+0i, 3+4i)
```

Using the **as.** functions, you can force coercion. Non-sensical coercion causes NAs - letters cannot be made into numbers, or logicals

```{r}
class(a)
a1 <- as.numeric(a) ## make all elements numbers (no change)
a2 <- as.logical(a) ## make all elements logical (T/F)
a3 <- as.character(a) ## make all elements characters
```

The **seq()** function allows you to vary the intervals in a vector, or specify a specific amount of of values between two numbers.

```{r}
b1 <- seq(from = 0, to = 10, by=0.5)
b2 <- seq(0, 10 , 0.5) ## without specification, seq assumes from, to and by
b3 <- seq(from = 3, to = 27, length.out = 9)
b4 <- seq(3, 27, length.out = 9)
```

## Lists

Lists can contain elements of different classes. Lists will have double brackets instead of single brackets when printed. Use **list()** to create.

```{r}
x <- list(1, "a", T, 1 + 4i)
```

## Matrices

Matrices, **matrix()**, are vectors with a dimension attribute (which is itself a vector containing two quantities; the number of rows and the number of columns). **dim() and attributes()** provide the dimension attribute and all attributes, respectively. Note that matrices fill column-wise.

```{r}
m <- matrix(1:12, nrow = 3, ncol = 4) ## creates a matrix with 4 rows and 3 columns, using the numbers 1 through 12. matrix is empty unless values specified.
dim(m) ## gives dimension attribute
attributes(m) # provides the attributes
print(m)
```

Alternatively, matrices can be made by applying specific dimensions to a given vector

```{r}
j <- 1:10 ## create a vector with integers 1 through 10
dim(j) <- c(2,5) ## change number of dimensions to 2 (nrows=2, ncolumns=5)
```

Also, you can use column binding, **cbind()**, and row binding, **rbind()**, to build matrices from vectors with compatible lengths

```{r}
x1 <- 1:5
x2 <- 121:125
x3 <- cbind(x1, x2)
x4 <- rbind(x1,x2)
```

# Data Types

## Factors

Factors are used to represent categorical data: either unordered (i.e. male or female) or ordered (ranked, like university positions). Essentially integer vectors where the vectors are linked to labels.These are treated differently by important modeling functions like **lm() and glm()**.

```{r}
f1 <- factor(c("yes", "no", "yes", "yes", "no", "yes", "yes", "maybe")) 
levels(f1) = c("yes", "no", "maybe") ## this assigns levels (yes 1st, no 2nd, maybe 3rd)
table(f1) ## gives frequency count each level in a factor
unclass(f1) ## will provide the numerical quantities underlying the levels, as well as order them according to this underlying quantity (i.e. whatever is "2" is printed second, the default order is alphabetical for different factors).
```

## Missing Values

NaN is given for improper/missing mathematical operations while NA is used for everything else. **is.na() and is.nan()** tests to see if objects are NA or NaN. NA values can have class (integer, character, etc.) while NaNs are numerical. NaN is a type of NA, NA is not a type of Nan.

```{r}
L <- c(1, 2, NA, 7, NaN, 10, 3)
is.na(L)
is.nan(L)
```

## Data Frame

Data frames are used to store tabular data. **data.frame()** can be used to generate one. They are a list where every element of the list has to have the same length; each element is essentially a column, and the length of the elements is essentially the rows. Basically what a list is to a vector, a data frame is to a matrix.

- these have special attributes **row.names** gives you the name of each row in a data set
- the **read.table() or read.csv()** commands will create a data frame for you
- can be converted to a matrix with **data.matric()**, this will coerce non-uniform elements. [**nrow(), ncol(), and dim() can provide dimensions**] 

```{r}
df1 <- data.frame(foo = 1:4, bar = c(T, T, F, T))
df1
```
## Names Attribute

R objects can be given names using **names()** function

```{r}
## 
N <- 1:10 ## create a vector
names(N)
names(N) <- c("A", "B", "C", "D", "E", "F","G", "H", "I", "J") ## name each column of the vector
N
```

```{r}
lis <- list(a = 1, b = "add", c = T) ## you can directly name lists as you create them
lis2 <- list(1, "add", T) ## or create the list first
names(lis2) <- c("a", "b", "c") ## then add the names
```

Use **dimnames()** when giving matrices names

```{r}
NB <- matrix(1:8, nrow = 4, ncol = 2) ## a matrix, 1 to 8, 4 rows and 2 columns
dimnames(NB) <- list(c("a", "b", "c", "d"), c("1", "2")) # first name the rows, a-d, then name the columns, 1 and 2.
NB
```

# Reading Tabular Data

**read.table and read.csv** will read text files that contain data stored in rows and columns, converting them to a data table. **readLines** reads lines of text files, converting text to character vector. **source** will read code into R, **dget** will read R code that has been dparsed into tex  **load and unserialize** functions are for reading binary objects into R.

All these have analogous functions for writing data into files:

- write.table
- writeLines
- dump
- dput
- save
- serialize

**read.table** has many arguments:
- file, name of the file or connection
- header, logical to indicate if file has header
- sep, string indicating column separation
- colClasses, character vector indicating class of each column
- comment.char, character string indicating comment character
- skip, the number of lins to skip at the beginning
- stringsAsFactors, should character variables be coded as factors? logical, defaults true

```{r}
```

**read.table** will skip lines that start with a #, figure out dimensions, figure out the type of variable in each column. **read.csv** is identical, but works on matrices separated by commas, not spaces. 

# Reading Large Datasets

Quick tips:

- read the help page for read.table
- make a rough calculation of the memory (RAM) needed for your dataset
- set comment.char = "" if there are no commented lines in your file
- use the **colClasses** argument to let R know what type of data is in each column (logical, numeric, integer, complex, character, raw). if you give a single class, R assumes they are all of that classes.

Below is a "quick and dirty" way to determine what types of data are in each column if you do not know/have too many columns. Note, **sapply()** can be used to apply a function over a list or vector. Note, if choosing a small subset of data to read, be sure it includes all possible types of data that column contains so R can do the appropriate coercion  - otherwise error ensues when column classes for the  whole data set are incorrectly labeled:

```{r}
initial <- read.table("AnonymousCovidData.csv", header = T, na.strings = c(""," ","NA"), sep =",", skip = 1, nrows = 1000, fill = T) ## import the table, specify there is a header, consider all empty cells or cells with " " to be NA, skip the first row, fill in empty cells and specify a small subset of 1000 rows to read through
classes <- sapply(initial, class) ## apply the class function to each and every element (column) of the data table subset
tabAll <- read.table("AnonymousCovidData.csv", header = T, na.strings = c(""," ","NA"), sep = ",", skip = 1, colClasses = classes, nrows = 4000, fill = T) # import table again, this time using all the data, and classify columns according to what R determined
```

# Textual Data Formats

**dump** and **dput** are useful because they can contain metadata on the actual data set. Textual formats are also more easily read by version control software, and corruption is easier to root out. Unfortunately, this format is not space efficient.

```{r}
y <- data.frame(a = 1, b = "a")
dput(y, file = "y.R")
y2 <- dget("y.R")
```

the **dput** function takes R objects and restructures it. You can later use **dget** to reconstruct the object as it was in its original format.

**dump** and **source** work similarly but can be applied to multiple R data structures simultaneously:

```{r}
t <- data.frame(time = seq(0, 20, by = 5), temp = c(12, 20, 27, 30, 25))
u <- "We the People"
v <- F
dump(c("t", "u", "v"), file = "agg.R") ## create an R structure
rm("t", "u", "v") ## use rm() function to remove created variables from environment
source("agg.R") # use source to bring back the removed variables via the "agg.R" structure
```

# Interfaces with the Outside World

Certain functions open up "connections" to the outside world. 

- **file** opens a connection to a file
- **gzfile** and **bzfile** open up connections to compressed files with gzip and bzip2, respectively
- **url** opens up a webpage
- **str** can be used to describe the arguments of a function: 

File: Open argument inputs include **r** for read only, **w** for write (initalize new file) and **a** for appending. **rb, wb, ab** for reading, writing and appending in binary

```{r}
str(file) ## description argument is name of code, open argument has different codes
```


```{r}
con <- file("Random_txt_file.txt", "r")
data <- read.csv(con)
close(con)
```
is the same as
```{r}
read.csv("Random_txt_file.txt")
```

Connections are more useful for reading parts of a file. Using **file**, or the appropriate derivative, you can open a file but use **readLines** to look at just part of it. A similar principle can be applied to web pages as well:

```{r}
con <- file("Random_txt_file.txt")
s <- readLines(con, 4)
```

```{r}
con <- url("https://www.jhsph.edu", "r")
u <- readLines(con)
head(u) ## returns first parts of a vector
tail(u) ## returns last parts of a  vector
close(con) ## remember to close connections when done
```


# Subsetting in R

There are different ways to extract subsets:

- **[** returns an object of the same class as the original, can be used to select more than one element (with one exception)
- **[[** extracts elements of a list or data frame. It can only take one element and the class of the extracted list is not certain to be the same
- **$** is to extract from a list or data frame by name

```{r}
d <- c("a","g", "b", "c", "z", "d", "e", "f")
x <- list(1, "a", T, 1 + 4i)
m <- matrix(1:12, nrow = 3, ncol = 4)

d[6]
d[3:7]
d[d < "d"]
u <- d < "e" ## logical vector that determines if each element is greater or less than a certain value
u
d[u] ## subset the full list with the logical vector to return only the desired values
```

Extracting a single element from a list:

```{r}
d <- c("a","g", "b", "c", "z", "d", "e", "f")
x <- list(1, "a", T, 1 + 4i)
m <- matrix(1:12, nrow = 3, ncol = 4)
o <- list(letters = d, list = x, matrix = m)

y1a <- o[3] ## returns data in same class as original. Meaning it extracts the matrix as the (only) item of a list, since o is a list
y1b <- o[c(1,3)] ## to return multiple outputs, a single bracket operator must be used, double brackets does not support multiple elements

y2 <- o[[3]] ## returns data of any class so takes out matrix as a matrix

y3a <- o$letters ## using the dollar sign, I asked for the list associated with letters
name <- "letters"
y3b <- o[["letters"]] ## does the same thing as o$letters
y3c <- o["letters"] ## will give  a list with the first element being the vector I asked for
y3d <- o[[name]]
```

Important to note, the **[[** operator can be used with computed indices, while the **$** operator needs literal names

Subsetting a matrix brings up the exception to single bracket subsetting. Instead of a matrix with one value, it returns a vector of length one with that single value.

```{r}
d <- c("a","g", "b", "c", "z", "d", "e", "f")
x <- list(1, "a", T, 1 + 4i)
m <- matrix(1:12, nrow = 3, ncol = 4)
o <- list(letters = d, list = x, matrix = m)

z1a <- o[[c(2,3)]] ## double subset to extract specific item
z1b <- o[[2]][[3]] ## another means of double subsetting

z2a <- m[1,2] ## will give the first row second column of a matrix
z2b <- m[1,]  ## will give the first row of the matrix
z2c <- m[,2] ## will give the second column of a matrix
z2d <- m[1,2, drop = F] ## causes the single bracket to keep (i.e. not drop) the second dimension of the matrix. Providing a 1x1 matrix with the value desired.
z2e <- m[1, , drop =F] ## similarly, whole rows/columns can be retained as 1xn or nx1 matrices.
```

## Partial Matching

Double bracket and dollar sign operator compatible. 

```{r}
d <- c("a","g", "b", "c", "z", "d", "e", "f")
x <- list(1, "a", T, 1 + 4i)
m <- matrix(1:12, nrow = 3, ncol = 4)
o <- list(letters = d, list = x, matrix = m)

o$le ## partial matching will "finish" the name for me, so assumes "le" is for "letters" since its the only match
o[["le", exact = FALSE]] ## the same thing can be achieved with double brackets but you need to set exact = false.
```

## Removing NA values

Exclamation piint, **!**, is bang operator, takes the opposite.

```{r}
k <- c(1, 2, NA, 4, NA, 5)
j <- c("a", "b", NA, "d", NA, "f")
bad <- is.na(k) ## create logical vector that confirms NA or not
noNA <- k[!bad] ## the elements of k that do NOT have NA (F with logical vector)
theNA <- k[bad] ## the elements of k that DO have NA (T with logical vector)
good <- complete.cases(k, j)
k[good]
j[good]
```

```{r}
initial <- read.table("AnonymousCovidData.csv", header = T, na.strings = c(""," ","NA"), sep =",", skip = 1, nrows = 1000, fill = T) ## import the table, specify there is a header, consider all empty cells or cells with " " to be NA, skip the first row, fill in empty cells and specify a small subset of 1000 rows to read through
classes <- sapply(initial, class) ## apply the class function to each and every element (column) of the data table subset
tabAll <- read.table("AnonymousCovidData.csv", header = T, na.strings = c(""," ","NA"), sep = ",", skip = 1, colClasses = classes, nrows = 4000, fill = T) # import table again, this time using all the data, and classify columns according to what R determined

tabAll[1:6, 1:5] ## import the first 6 rows and first five columns of this data set
good <- complete.cases(tabAll)
## tabAll[good][1:6,1:5] ## wont work for me because absolutely none of rows are completely filled in
```

# Vectorized Operations

Some processes are done in parallel (applied across all elements of a vector):

```{r}
j <- 1:10
k <- 11:20
j + k ## element-wise sum of vectors
j <= 6 ## logical vector denoting which elements satisfy condition
j * k ## element wise product of vectors
```


```{r}
m1 <- matrix(1:4, 2, 2); m2 <- matrix(rep(10,4), 2, 2)
m1 * m2 ## this performs the dot product (element-wise) product of the matrices
m2 / m1
m1 %*% m2 ## surround in percentage signs to ensure matrix multiplication
```

# HW1

```{r}
HW1 <- read.csv("hw1_data.csv")
HW1[1:2,] ## print first two rows
nrow(HW1) ## calculate the number of rows
tail(HW1, n = 2L) ## print the last two rows of the data sheet
HW1[47, 1]
Oz <- HW1[,1] ## create a vector containing all the data from the ozone column
is.na(Oz) ## create a logical vector denoting which cells have NA
length(Oz[is.na(Oz)]) ## apply the logical vector to the whole column, returning only the columns with values, then calculate the length of that vector
mean(Oz[!is.na(Oz)]) ## average the the non-NA values of the vector using bang operator and mean function
trunc <- HW1[HW1$Ozone > 31 & HW1$Temp >90,] ## create a truncated table that removes columns that have ozone values above 31 and temp values above 90
## sol1 <- is.na(trunc$Solar.R)
sol <- trunc$Solar.R[!is.na(trunc$Solar.R)] ## create a vector with the values from the truncated solar.R column that excludes NA
mean(sol) ## mean of values in this range
M6 <- HW1[HW1$Month == 6,] ## retain only values for month 6
mean(M6$Temp)
M5 <- HW1[HW1$Month == 5,]  ## retain only values for month 5 in table
M5b <- M5$Ozone[!is.na(M5$Ozone)] ## create a vector of the non-NA values of the ozone column in the month 5 subset
max(M5b)
```